The Galactic Midplane Is Not a Plane: Implications for Dynamical Analysis with \textit{Gaia} Data and Beyond

### 2019-05-09 paper/ms.tex
moved discussion of measurement uncertainties to discussion, and added some to conclusions list

\section{Coordinate System Offsets I. Measurement Errors} \label{sec:mes_err}
We consider the first potential source of coordinate system offsets which
would generate the action errors described in Section~\ref{sec:ref_frame}:
uncertainties in measurements of the position and velocity of the Sun relative
to the Galactic center. Converting from heliocentric to Galactocentric
coordinates relies on these measurements. Therefore, errors in their values
will induce a systematic offset in the Galactocentric phase-space position of
any observed star.

\subsection{Galactic Center}
First, one must define the center of the Galaxy. This is usually taken to be
the location of the central supermassive black hole, Sagittarius~A\textsuperscript{*}
\citep[\sgra{}, e.g.][]{2004ApJ...616..872R}. From stellar motions near
\sgra{}, the distance from the Sun to \sgra{}, referred to as $R_0$, can be
precisely measured \citep{2009ApJ...692.1075G, 2018AA...615L..15G}, with a
recent measurement using near-infrared interferometry of $8.178 \pm 0.035$~kpc
\citep{2019arXiv190405721A}.

\subsection{Galactic Orientation}
Second, one must define the angular orientation of the Galaxy. This was
defined in 1958 by the IAU subcomission 33b \citep{1960MNRAS.121..123B} by
defining the coordinates of the Galactic center in B1950 coordinates as
(17:42:26.6, -28:55:00) and the North Galactic pole as (12:49:00, +27:24:00).
These two quantities, together with $R_0$, define the orientation of the
Galactic plane. However, there is growing evidence that the stellar midplane
is tilted relative to this coordinate system \citep{2014ApJ...797...53G,
2016ARAA..54..529B}, though not the H~\textsc{ii} midplane
\citep{2019ApJ...871..145A}.

\subsection{Solar Height}
Third, one must define the Sun's vertical distance from the Galactic midplane,
which can be determined by identifying where the stellar density and
velocities reach a maximum (effectively the median height of all disk stars).
The Solar height is usually taken to be $\sim 25\,\pc$
\citep{2001ApJ...553..184C}, with a more recent measurement from \textit{Gaia}
DR2 placing it at $20.8 \pm 0.3\,\pc$ \citep{2019MNRAS.482.1417B}. Another
strategy is to use the cold gas or H~\textsc{ii} regions in the disk to define
the Galactic midplane, leading to slightly different values (by $\sim 5\,\pc$)
for the Sun's relative height \citep[e.g.][]{2019ApJ...871..145A}. A pre-{\em
Gaia} review of these measurements is given by \citet{2016ARAA..54..529B}.

\subsection{Local Standard of Rest}
Finally, one must define the LSR, or mean velocity of stars near the Sun
relative to the Galactic center (which is defined to have zero velocity), and
the velocity of the Sun relative to the LSR. The radial ($U_{\odot}$) and
vertical ($W_{\odot}$) components are computed by taking the mean motions of
different stellar groups \citep[e.g.][]{2012MNRAS.427..274S}. The azimuthal
component ($V_{\odot}$) is more difficult to measure, but can be modeled using
the asymmetric drift relation \citep{2008gady.book.....B}. The values of the
components of the LSR are usually taken from \citet{2010MNRAS.403.1829S}. The
value of the circular velocity is taken to be $\sim 220\kms$
\citep[e.g.][]{2012ApJ...759..131B}.

Considerable effort has been placed on each of these measurements, but
uncertainties remain, and detailed dynamical modeling across the disk ---
particularly for dynamically cold stars --- may have to take them into
account. For instance, the discrepancy of $\sim 5\,\pc$ between the midplane
defined by stars and by gas will induce a $\sim10\%$ error in $J_z$ for an
orbit with $z_{\text{max}}\sim100\,\pc$ (see Section~\ref{sec:ref_frame}).

@@

Aside from measurement uncertainties, intrinsic variations of the midplane position across the disk of the galaxy can contribute to systematic errors in the actions computed assuming an axisymmetric potential model. The midplane position can vary as a function of location due to small local variations in the density of stars and gas in the Galaxy. Hints of this variation as a function of Galactocentric radius have been noted through their impact on the stellar velocity distribution by \citet{2012ApJ...750L..41W},\citet{2013ApJ...777L...5C}, and \citet{2013MNRAS.436..101W} pre-Gaia and recently by \citet{2019arXiv190209569F} post-Gaia. As pointed out by e.g. \citet{2014ApJ...797...53G} and \citet{2019ApJ...871..145A}, among many others, the gas distribution in the Galaxy also shows significant density variation across the disk. In this work, to stay close to the assumptions of our analytic calculations in \S \ref{ssec:epi_action}, we specifically consider \emph{azimuthal} variations in the midplane at the Solar circle, as defined by the stellar mass density. Since to our knowledge there are no direct empirical measurements of these variations for the Milky Way, we use two sets of simulations to estimate the size of this effect. One set are three zoomed, cosmological-hydrodynamical simulations of Milky Way-mass galaxies from the FIRE collaboration, described briefly in \S \ref{ssec:cosmozoom}, which include stars, gas, and dark matter in a fully cosmological setting but are not tailored to specific properties of the MW (such as the scale height or scale length, or the details of the accretion history). We use these simulations to span the range of possibilities for azimuthal midplane variations. The other simulations are isolated N-body simulations of interactions between the MW and a Sagittarius-like dwarf galaxy, described briefly in \S \ref{ssec:sag_sim}; these include dark matter and stars and are tailored to existing measurements of the structure of the MW's disk and of the orbit and properties of the Sagittarius (Sgr) dwarf galaxy. Comparing the azimuthal midplane variations in these simulations before and after the interaction with Sgr gives an idea of the effect of one minor merger whose properties are relatively well measured.

@@

Cosmological simulations of Milky Way-mass galaxies are not perfect
representations of the true Milky Way in other ways as well,\footnote{The
failure of cosmological simulations to exactly reproduce the Milky Way is not
entirely due to limitations of the numerical model. Candidate Milky Way-like
galaxies are chosen solely on their mass and isolation, for which there are a
wide variety of possible galaxies with qualitatively different properties.} as
discussed in \citet{2018arXiv180610564S}. For instance, the velocity structure
of \mi{} is closer to M31's than the Milky Way's (Loebman et al. in prep).
However, in this work we are most interested in the global properties of the
potential, and specifically in deviations from axisymmetry. From this
perspective, the simulated galaxies are actually \emph{more} axisymmetric than
we might expect of the Milky Way. While they have prominent spiral arms, none
has as strong a bar as the Milky Way does at present day, and none has a
nearby companion like the Large Magellanic Cloud. One of the three we consider
(\mf) does have an ongoing interaction with a satellite galaxy similar to
Sagittarius, which has punched through the galactic disk outside the solar
circle and induced some warping.

@@

\section{Discussion} \label{sec:discussion} 
We have used high-resolution cosmological simulations to illustrate that we
expect the ``local midplane'' defined by stellar density to vary with azimuth
by up to $\pm 100$ pc, as a natural consequence of the non-axisymmetry of the
Galactic disk at small scales.

While this is not in itself surprising or new, we also demonstrated that
extending the coordinate system established by our local midplane to a global
axisymmetric coordinate system spanning the entirety of the Galactic disk
introduces systematic error in computations of the actions under this symmetry
assumption, when starting from the present-day positions and velocities of
stars as measured by, e.g., \emph{Gaia}.

@@

\subsection{Estimate of Milky Way Midplane Offsets from Current Data}
\label{ssec:mw_data_midplane}

@@

\subsection{Coordinate System Offsets}\label{ssec:coord_off}
In Section~\ref{sec:mes_err} we discussed the current observational efforts to
measure the parameters of the Galactocentric coordinate system. We noted that
the solar height is discrepant between gas-based and stellar-based
determinations by $\sim5\,\pc$. While the effect is small and thus only likely
to dominate the intrinsic midplane variations on small scales, it may be
important for detailed modeling of kinematically cold stars where a $5\,\pc$
offset is relevant.

We also noted that the distance from the Sun to \sgra{} has been recently
measured to a very low uncertainty of $\sim0.3\%$ or $\pm25\,\pc$
\citep{2019arXiv190405721A}. However, the location of \sgra{} may not be
equivalent to the location of the ``dynamical Galactic center,'' i.e., the
point in three-dimensional space about which the stars in the solar
neighborhood are orbiting. This assumption, although sensible and frequently
made, has not yet been justified.

@@

@@

\item To demonstrate the previous point, we measured the local galactic
midplane along the Solar circle in three different high-resolution, zoom-in
simulations of Milky Way mass galaxies from the FIRE collaboration. We found
that the midplane varies as a function of azimuth at the Solar circle by
$\sim185, 162, 84\,\pc$ (middle $90\%$) for the three galaxies we considered
(\mi{}, \mf{}, and \mm{}, respectively).

@@

\item While inaccuracies in the parameters of the currently adopted
Galactocentric coordinate system are likely to be unimportant for most
applications, we discussed the importance of testing the assumption that the
dynamical Galactic center is colocated with \sgra{}, which we mentioned how to
do.

>>>>

@@

Intrinsic variations of the midplane position across the disk of the galaxy can contribute to systematic errors in the actions computed assuming an axisymmetric potential model. The midplane position can vary as a function of location due to small local variations in the density of stars and gas in the Galaxy. Hints of this variation as a function of Galactocentric radius have been noted through their impact on the stellar velocity distribution by \citet{2012ApJ...750L..41W},\citet{2013ApJ...777L...5C}, and \citet{2013MNRAS.436..101W} pre-Gaia and recently by \citet{2019arXiv190209569F} post-Gaia. As pointed out by e.g. \citet{2014ApJ...797...53G} and \citet{2019ApJ...871..145A}, among many others, the gas distribution in the Galaxy also shows significant density variation across the disk. In this work, to stay close to the assumptions of our analytic calculations in \S \ref{ssec:epi_action}, we specifically consider \emph{azimuthal} variations in the midplane at the Solar circle, as defined by the stellar mass density. Since to our knowledge there are no direct empirical measurements of these variations for the Milky Way, we use two sets of simulations to estimate the size of this effect. One set are three zoomed, cosmological-hydrodynamical simulations of Milky Way-mass galaxies from the FIRE collaboration, described briefly in \S \ref{ssec:cosmozoom}, which include stars, gas, and dark matter in a fully cosmological setting but are not tailored to specific properties of the MW (such as the scale height or scale length, or the details of the accretion history). We use these simulations to span the range of possibilities for azimuthal midplane variations. The other simulations are isolated N-body simulations of interactions between the MW and a Sagittarius-like dwarf galaxy, described briefly in \S \ref{ssec:sag_sim}; these include dark matter and stars and are tailored to existing measurements of the structure of the MW's disk and of the orbit and properties of the Sagittarius (Sgr) dwarf galaxy. Comparing the azimuthal midplane variations in these simulations before and after the interaction with Sgr gives an idea of the effect of one minor merger whose properties are relatively well measured.

@@

Cosmological simulations of Milky Way-mass galaxies are not perfect
representations of the true Milky Way in other ways as well,\footnote{The
failure of cosmological simulations to exactly reproduce the Milky Way is not
entirely due to limitations of the numerical model. Candidate Milky Way-like
galaxies are chosen solely on their mass and isolation, for which there are a
wide variety of possible galaxies with qualitatively different properties.} as
discussed in \citet{2018arXiv180610564S}. For instance, the velocity structure
of \mi{} is closer to M31's than the Milky Way's (Loebman et al. in prep).
However, in this work we are most interested in the global properties of the
potential, and specifically in deviations from axisymmetry. From this
perspective, the simulated galaxies are actually \emph{more} axisymmetric than
we might expect of the Milky Way. While they have prominent spiral arms, none
has as strong a bar as the Milky Way does at present day, and none has a
nearby companion like the Large Magellanic Cloud. One of the three we consider
(\mf) does have an ongoing interaction with a satellite galaxy similar in mass to
Sagittarius, which has punched through the galactic disk outside the solar
circle, leaving behind some of its stars and inducing warping in the disk.

@@

\section{Discussion} \label{sec:discussion} 
We have used high-resolution simulations to illustrate that we
expect the ``local midplane'' defined by stellar density to vary with azimuth
by up to $\pm 100$ pc, as a natural consequence of the non-axisymmetry of the
Galactic disk at small scales. While this is not in itself surprising or new, we also demonstrated that extending the coordinate system established by our local midplane to a global axisymmetric coordinate system spanning the entirety of the Galactic disk
introduces systematic error in computations of the actions under this symmetry
assumption, when starting from the present-day positions and velocities of
stars as measured by, e.g., \emph{Gaia}.

@@

\subsection{Estimates of Milky Way Midplane Offsets from Current Data}
\label{ssec:mw_data_midplane}

@@

\subsection{Uncertainties in the Solar position and velocity}\label{ssec:coord_off}

Uncertainties in measurements of the position and velocity of the Sun relative
to the Galactic center can also contribute to systematic error in the actions, since converting from heliocentric to Galactocentric
coordinates relies on these measurements. Therefore, errors in their values
will induce a systematic offset in the Galactocentric phase-space position of
any observed star. Considerable effort has been placed on each of these measurements, but uncertainties remain, and detailed dynamical modeling across the disk---particularly for dynamically cold stars---may have to take them into
account. 
Here we briefly review the current measurements and their uncertainties, and the implications for the calculation of actions.

\subsubsection{Galactic Center position}
First, one must define the center of the Galaxy. This is usually taken to be
the location of the central supermassive black hole, Sagittarius~A\textsuperscript{*}
\citep[\sgra{}, e.g.][]{2004ApJ...616..872R}. From stellar motions near
\sgra{}, the distance from the Sun to \sgra{}, referred to as $R_0$, can be
precisely measured \citep{2009ApJ...692.1075G, 2018AA...615L..15G}, with a
recent measurement using near-infrared interferometry of $8.178 \pm 0.035$~kpc
\citep{2019arXiv190405721A}, or $\sim0.3\%$. However, the location of \sgra{} may not be equivalent to the location of the ``dynamical Galactic center,'' i.e., the
point in three-dimensional space about which the stars in the solar
neighborhood are orbiting. This assumption, although sensible and frequently
made, has not yet been justified.

@@

\subsubsection{Galactic Orientation}
Second, one must define the angular orientation of the Galaxy. This was
defined in 1958 by the IAU subcomission 33b \citep{1960MNRAS.121..123B} by
defining the coordinates of the Galactic center in B1950 coordinates as
(17:42:26.6, -28:55:00) and the North Galactic pole as (12:49:00, +27:24:00).
These two quantities, together with $R_0$, define the orientation of the
Galactic plane. However, there is growing evidence that the stellar midplane
is tilted relative to this coordinate system \citep{2014ApJ...797...53G,
2016ARAA..54..529B}, though not the H~\textsc{ii} midplane
\citep{2019ApJ...871..145A}.  This tilt will contribute a linear dependence in Galactocentric radius to the systematic errors in actions induced by midplane offsets.

\subsubsection{Solar Height}
Third, one must define the Sun's vertical distance from the Galactic midplane,
which can be determined by identifying where the stellar density and
velocities reach a maximum (effectively the median height of all disk stars).
The Solar height is usually taken to be $\sim 25\,\pc$
\citep{2001ApJ...553..184C}, with a more recent measurement from \textit{Gaia}
DR2 placing it at $20.8 \pm 0.3\,\pc$ \citep{2019MNRAS.482.1417B}. Another
strategy is to use the cold gas or H~\textsc{ii} regions in the disk to define
the Galactic midplane, leading to slightly different values (by $\sim 5\,\pc$)
for the Sun's relative height \citep[e.g.][]{2019ApJ...871..145A}. A pre-{\em
Gaia} review of these measurements is given by \citet{2016ARAA..54..529B}. The discrepancy between gas-based and stellar-based determinations of the solar height is small, and thus only likely to dominate over intrinsic midplane variations on small scales, but will be relevant for detailed modeling of young and kinematically cold stars. For instance, it will induce a $\sim10\%$ error in $J_z$ for an
orbit with $z_{\text{max}}\sim100\,\pc$ (see Section~\ref{sec:ref_frame}).

\subsubsection{Local Standard of Rest}
Finally, one must define the LSR, or mean velocity of stars near the Sun
relative to the Galactic center (which is defined to have zero velocity), and
the velocity of the Sun relative to the LSR. The radial ($U_{\odot}$) and
vertical ($W_{\odot}$) components are computed by taking the mean motions of
different stellar groups \citep[e.g.][]{2012MNRAS.427..274S}. The azimuthal
component ($V_{\odot}$) is more difficult to measure, but can be modeled using
the asymmetric drift relation \citep{2008gady.book.....B}. The values of the
components of the LSR are usually taken from \citet{2010MNRAS.403.1829S}. Their uncertainties should also lead to systematic errors in the actions, as given in Eqs. \ref{eq:induced_Jphi}--\ref{eq:induced_Jz}. For example, the
value of the circular velocity is taken to be $\sim 220\kms$
\citep[e.g.][]{2012ApJ...759..131B} with roughly 10\% uncertainty. Based on the analytic estimates in \S \ref{ssec:epi_action}, we expect this to translate to at least a 10\% systematic error in $J_{\phi}$.

@@

\item To demonstrate the previous point, we measured the local galactic
midplane along the Solar circle in three different high-resolution, zoom-in
simulations of Milky Way mass galaxies from the FIRE collaboration, as well as a controlled simulation of the interaction of the MW with Sagittarius. We found
that the midplane varies as a function of azimuth at the Solar circle by
$\sim185, 162, 84\,\pc$ (middle $90\%$) for the three cosmological simulations we considered
(\mi{}, \mf{}, and \mm{}, respectively) and by $\sim30$ and $\sim100$ pc before and after the interaction in the controlled simulation. Comparison of the corresponding velocity variations with recent measurements in the MW suggests that these values are consistent with expectations for midplane variations in the MW itself, which have not yet been measured as a function of azimuth.

@@

\item Some inaccuracies in the parameters of the currently adopted
Galactocentric coordinate system are likely important for some
applications. In particular, it is imperative to test the assumption that the
dynamical Galactic center is colocated with \sgra{}. We discuss how to
do this in \S \ref{ssec:coord_off}.
